---
title: "Practical 4 - Hierarchical ascending classification"
author: "Agnès LAGNOUX & Nicolas SAVY"
date: ""
output: 
  pdf_document:
    highlight: tango  # Vous pouvez tester d'autres thèmes : kate, monochrome, etc.
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_chunk$set(fig.width = 6, fig.height = 4, out.width = "75%")
```

# Exercise 2

The goal is to find with R the results of Exercise 1.

1. Create a matrix X containing the data.

```{r}
X <- matrix(c(5,4,4,5,1,-2,0,-3),4,2,byrow=TRUE)
colnames(X) <- c("X1","X2")
rownames(X) <- c("1", "2","3","4")
```

2.  Visualize the dataset.

```{r out.width="75%"}
# Loading of ggplot2
library(ggplot2)

# Generation of the data
X <- data.frame(
  X1 = c(5, 4, 1, 0),
  X2 = c(4, 5, -2, -3),
  label = factor(1:4)
)

# Graphical reprezentation with ggplot2
ggplot(X, aes(x = X1, y = X2)) +
  geom_point(size = 4) +  # Points
  geom_text(aes(label = label), vjust = -0.5, hjust = 1.5) +  # Add the labels
  geom_hline(yintercept = 0, linetype = "dashed") +           # Horizontal line
  geom_vline(xintercept = 0, linetype = "dashed") +           # Vertical line
  theme_minimal() +                                           # Minimal theme
  xlim(-3.5, 5.5) +                                           # x-axis limits
  ylim(-3.5, 5.5) +                                           # y-axis limits
  labs(title = "Points cloud", x = "X1", y = "X2")
```

3.  Construct the hierarchy with the maximal link.

```{r}
d <- dist(X)                          # compute the Euclidean distances between points
treeC <- hclust(d, method="complete") # all the results are stored in the object named tree
```

4.  Determiner the heights of the classes.

```{r}
treeC$height # give the height of the classes in the dendogram
```

5.  Represent the dendrogram of such hierarchy.

```{r out.width="75%"}
plot(treeC, main="Maximal link's HAC", xlab ="", sub="", hang=-1)
```

6.  Cut the dendrogram to get the partition in two classes of the maximal link's HAC.

```{r}
cutree(treeC, k=2 )  # allow to obtain the partition in two classes
# here, the two classes are C1={1,2} et C2={3,4}
```

7.  Construct now Ward's hiérarchy and retrieve the lengths of Exercise 1.

```{r}
treeW <- hclust(d, method="ward.D2")
sum(treeW$height)
```

NB1: in this exercise, the weights of the individuals are all equal to 1.

NB2: with this setting, we find that the sum of the heights of
the tree is worth 67, that is to say the total inertia (here the sum of the total squares) calculated by hand and found previously with the $k$-means.

8.  Represent the dendrogram of such hierarchy.

```{r out.width="75%"}
plot(treeW, main="Ward's HAC", xlab ="", sub="",hang=-1)
cutree(treeW,k = 2) # we find again the two classes C1={1,2} et C2={3,4}
```

9.  Find from the heights of this dendrogram:

- the total inertia,

- the inter-class inertia of the partition into two classes.

Deduce the share of the total variance explained by this partition.

```{r}
# Calculation of the share of inertia explained by Ward partitions

# Total inertia = sum of heights
Tot <- sum(treeW$height)

# Inter-class inertia = sum of the k-1 greatest heights (k=number of classes)
h <- sort(treeW$height, decreasing=TRUE)
K <- 2
B <- sum(h[1:K-1])

# Part of inertia explained by the partition = inter inertia/total inertia
B/Tot 
```
